### 2020-01-11 manuscript/manuscript.rmd
update previous analyses

The data matrix was constructed using a semi-automated approach. Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplications. Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors). The remaining gene trees with more than one sequence for any taxon were then manually inspected. If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded. If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded. Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix. Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships. Unstable taxa were identified with leaf stability indices [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix. This resulted in the 64-taxon matrix that is the focus of most of their analyses. Phylogenetic analyses were conducted under the F81+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].

@@

Philippe *et al.* 2009 [Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals. The phylogenetic analysis using Poisson-CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade. The authors concluded:
> The resulting phylogeny yields two significant conclusions reviving old views that have been challenged in the molecular era: (1) that the sponges (Porifera) are monophyletic and not paraphyletic as repeatedly proposed, thus undermining the idea that ancestral metazoans had a sponge-like body plan; (2) that the most likely position for the ctenophores is together with the cnidarians in a ‘‘coelenterate’’ clade.

@@

Hejnol *et al.* 2009 added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix. The orthology inference was largely simialr to Dunn et al. 2008. Two datsets were constructed, one with 844 genes and the other with 330 genes. Maximum likelihood analyses from both datasets strongly supported Ctenophora-sister.

@@

### Ryan *et al.* 2013
Ryan *et al.* 2013 sequenced the first ctenophore genome of *Mnemiopsis leidyi*. With the genome resources of *M. leidyi*, the authors constructed two phylogenomic datasets: a "Genome set" based on 13 animal genomes and a "EST Set" that also included 59 animals. They analyzed both matrices by site-homogeneous GTR+Gamma model and Poisson-CAT model with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology. The results strongly supported Ctenophora-sister in all datasets they analysed using site-homogeneous model. The Poisson + CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days.

@@

Moroz *et al.* 2014 sequenced the second ctenophore genome *Pleurobrachia bachei* to explore the phylogenetic relationship of Metazoa. All phylogenetic analyses strongly supported Ctenophora-sister with different taxon and gene sampling using WAG site-homogeneous model.

### Whelan et al. 2015
Wehlan *et al.* 2015 constructed a new phylogenomic dataset by eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g. long-branch attraction, compositional bias, fast evolving genes, etc.). Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (~ 1000 orthologs) and subsequetly removal of genes with too much missing data and potential paralogs.  The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups. All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister. CAT-GTR models only used in least saturated dataset 6 and 16 also strongly supported Ctenophora. Regarding the recovery of Ctenophora-sister, the authors concluded:
> Importantly, biases resulting from elevated compositional heterogeneity or elevated substitution rates are ruled out. Placement of ctenophores as
sister to all other animals, and sponge monophyly, are strongly supported under multiple analyses, herein.

@@

Chang *et al.* 2015 was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity. The authors constructed a dataset with 200 protein markers based on Philippe et al. 2011 with 51,940 amino acids and 77 taxa. Both site-heterogeneous Poisson-CAT and site-homogeneous GTR models strongly supported Ctenophora-sister.

@@

Pisani *et al.* 2015 reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013, Moroz2014 and Whelan2015 datasets. It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of animal-root position. The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction. They found strong support of porifera-sister when only include Choanozoa as outgroup in these datasets using site-heterogeous CAT model. Regarding the recovery of Porifera-sister, the authors concluded:
>Our results reinforce a traditional scenario for the evolution of complexity in animals, and indicate that inferences about the evolution of Metazoa
based on the Ctenophora-sister hypothesis are not supported by the currently available data.

@@

Feuda *et al.* didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets). It was the first study phylogenomic analysis using recoding strategy to explore animal-root position. The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, PartitionFinder)and site-heterogeneous (CAT-GTR) models with non-recoding and recoding datasets. The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with CAT-GTR models and strongly supported Porifera-sister hypothesis. Regarding the recovery of Porifera-sister, the authors concluded:

@@

### Whelan and Halalnych 2017a

Whelan *et al.* is the only study to evaluate performance of site-heterogeneous models and site-homogeneous model with data partitioning under the simulation framework. The simulation results suggested that Poisson+CAT model consistently performed worse than other models in simulation datasets. More importantly, the authors also showed that both Poisson + CAT and GTR +  CAT models could overestimated substitutional heterogeneity in many cases. They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site -homogeneious model. The results indicated that Poisson + CAT model tends to recover less accurate trees and both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses. The authors also concluded:

@@

### Whelan *et al.* 2017b

Whelan *et al.* added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015, with the difference of a core Ctenophora core datasets were constructed here with more than 2000 genes. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in Choanoazoa dataset. Regarding the recovery of Ctenophora-sister, the authors concluded:
>Using datasets with reasonably high ctenophore and other non-bilaterian taxon sampling, our results strongly reject the hypothesis that sponges are the sister lineage to all other extant metazoans.

@@

Simion *et al.* [@Dunn:2008ky] added transcriptomic data for 21 new animals. The data matrix was constructed using a semi-automated approach to comprehensively detect and eliminate potential systematic errors. The resulting dataset comprises 1,719 genes and 97 species, including 61 non-bilaterian metazoan species. It was by far the largest phylogenomic dataset in terms of taxon and gene sampling related to animal-root question.

The supermatrix was first analyzed using the Poisson-CAT model. Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size. Each jackknife is baed on a random selection of ~ 25% of the genes. The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in the supermatrix dataset. Importantly, the authors compared the behavior of long-branch effect between site-heterogenious and site-homogeneous models by progressively removal taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models. Regarding the recovery of Ctenophora-sister, the authors concluded:

@@

>>>>

The data matrix was constructed using a semi-automated approach. Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplications. Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors). The remaining gene trees with more than one sequence for any taxon were then manually inspected. If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded. If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded. Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix. Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships. Unstable taxa were identified with leaf stability indices [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix. This resulted in the 64-taxon matrix that is the focus of most of their analyses. Phylogenetic analyses were conducted under the Poisson+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].

@@

Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals by added 9 new species. Th data matrix were assembled based on The phylogenetic analysis using Poisson-CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade. Gene trees were then constructed, and potentially paralogs were removed by a bootstrap threshold of 70. Ambiguously aligned regions were trimemed and only genes sampled for at least two-thirds of species were retained. The phylogenetic analyses were conducted under the Poisson+CAT model in phylobayes [@Lartillot:2004dq].

> The resulting phylogeny yields two significant conclusions reviving old views that have been challenged in the molecular era: (1) that the sponges (Porifera) are monophyletic and not paraphyletic as repeatedly proposed, thus undermining the idea that ancestral metazoans had a sponge-like body plan; (2) that the most likely position for the ctenophores is together with the cnidarians in a ‘‘coelenterate’’ clade.

@@

Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatwroms. The orthology inference was largely similar to Dunn et al. 2008, with the exception of orthology genes were clustered by MCL. The final data matrix included 1497 genes, and then subsampled with 844, 330 and 53 gens by different thresholds of gene occupancy. With the exception of 53 gene matrix, maximum likelihood analyses from all other datasets strongly supported Ctenophora-sister (models were selected by RaxML perl script).

@@

### Nosenko *et al.* 2013
Nosenko *et al.* [@nosenko2013deep] added Expressed Sequence Tag (EST) data for 9 species of non-bilaterian metazoans (7 sponges). They constructed a novel matrix containing 122 genes and parsed them into two non-overlaping matrices (ribosomal and non-ribosomal genes) and found incongruent results of deep metazoan phylogeny. The other major finding was that ribosomal gene partitions showed  significantly lower saturation than the non-ribosomal ones.

OGs were constructed using the bioinfomatics pipeline OrthoSelect [@schreiber2009orthoselect]. They also evaluated level of saturation, leaf stability of sampled taxa, compositional heterogeinity and model comparison of each matrix. By modifying gene sampling, ingroup and outgroup sampling, three major topologies related to the position of animal-root were constructed (including Porifera+Placozoa sister, Ctenophora-sister and Porifera-sister). Phylogenetic analyses were conducted under the Poisson+CAT, GTR+CAT and GTR models in phylobayes [@Lartillot:2004dq].

> we were able to reconstruct a metazoan phylogeny that is consistent with traditional, morphology-based views on the phylogeny of non-bilaterian metazoans, including monophyletic Porifera and ctenophores as a sister-group of cnidarians.

Note the main conclusion of this paper is baed on the selection of genes that evolved slowly ribosomal RNA genes. The Ctenophora-sister was rejected based on the most saturated dataset and not supported by the matrix with the most closely related outgroups (Choanomalia matrix) [@nosenko2013deep].

### Ryan *et al.* 2013
Ryan *et al.* sequenced the first ctenophore genome of *Mnemiopsis leidyi*. With the genome resources of *M. leidyi*, the authors constructed two phylogenomic datasets: a "Genome set" based on 13 animal genomes and a "EST Set" that also included 59 animals. They analyzed both matrices by site-homogeneous GTR+Gamma and site-heterogeneous Poisson-CAT models with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology for the Ryan2013_est matrix. The OGs were constructed based on the method of Hejnol *et al.* 2009. For the Ryan2013_genome matrix, they performed phylogenetic analyses with both gene content and sequence-baed analyses. Overall, their results strongly supported Ctenophora-sister in all datasets they analyzed using site-homogeneous model. The Poisson+CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days (but strongly supported Porifera in Choaimalia matrix).

>Our phylogenetic analyses suggest that ctenophores are the sister group to the rest of the extant animals.

@@

Moroz *et al.* [@moroz2014ctenophore] sequenced the second ctenophore genome *Pleurobrachia bachei* to explore the phylogenetic relationship of Metazoa. All phylogenetic analyses strongly supported Ctenophora-sister with different taxon and gene sampling using WAG site-homogeneous model. Two phylogenomic matrices were generated, the first set was represented by two ctenophore species, whereas the other set contained improved ctenophore sampling (10 taxa, Moroz2013_3d). Orthology determination employed in HaMStR [@ebersberger2009hamstr] using 1,032 "model organism" single-copy OGs. Sequence were then trimmed and aligned. This resulted in a final matrix of 170,871 amino acid positions across 586 genes with 44 taxa for the first matrix, and 114 genes with 60 taxa for the second matrix. All the phylogenetic analyses were analyzed in RAxML under the WAG+CAT+F models (different than CAT models in PhyloBayes) to reduce the computational cost.

>Our integrative analyses place Ctenophora as the earliest lineage within Metazoa. This hypothesis is supported by comparative analysis of multiple gene families, including the apparent absence of HOX genes, canonical microRNA machinery, and reduced immune complement in ctenophores.

It should be noted that only the Moroz_3d matrix has been reanalyzed in other studies, although the support of Ctenophora-sister is quite low.

### Borowiec *et al.* 2015

Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprised 1080 OGs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included. The OGS were constructed using OrthologID pipeline [@chiu2006orthologid]. After removal of spurious sequences and genes with more than 40% of mission data, the final matrix included 1080 OGs (Total 1080). The authors further filtered the full dataset to 9 sub-datasets by filtering genes with high long-branch scores; genes with high saturation; gene occupancy; fast evolving genes. The main conclusion of the paper was largely based on BorowiecTotal_1080 and Borowiec_Best108 matrices. Phylogenetic analyses were conducted under the GTR+CAT model in PhyloBayes in selected matrices, and under the data-partitioning methods in RAxML for all matrices.

>Our phylogeny
supports the still-controversial position of ctenophores as sister group to all other metazoans. This study also
provides a workflow and computational tools for minimizing systematic bias in genome-based phylogenetic
analyses.

It should be noted that the authors also employed recoding-method in the Borowiec_Best108 matrix and found neither support of Porifera-sister or Ctenophora-sister [@borowiec2015extracting].

### Whelan et al. 2015
Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic dataset by eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g. long-branch attraction, compositional bias, fast evolving genes, etc.). Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.  The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups. All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister. CAT-GTR models only used in least saturated dataset 6 and 16 also strongly supported Ctenophora.

> Importantly, biases resulting from elevated compositional heterogeneity or elevated substitution rates are ruled out. Placement of ctenophores as
sister to all other animals, and sponge monophyly, are strongly supported under multiple analyses, herein.

Note that the authors also reanalyzed Philippe2009 matrix (with the removal of ribosomal genes) and recovered Porifera-sister with moderate support (pp=90).

@@

Chang *et al.* [@Chang:2015hl] was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity. The authors constructed a dataset with 200 protein markers based on Philippe et al. 2011 [@philippe2011resolving] with 51,940 amino acids and 77 taxa. Both site-heterogeneous Poisson-CAT and site-homogeneous GTR models strongly supported Ctenophora-sister.

Note that this data matrix has been extensively reanalyzed in many studies.

@@

Pisani *et al.* [@pisani2015genomic] reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013_est, Moroz2014_3d and Whelan2015 datasets. It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of the position of the root of animal phylogeny. The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction. Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in phylobayes. They found GTR-CAT and Poisson-CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice. Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.

Regarding the recovery of Porifera-sister, the authors concluded:

>Our results reinforce a traditional scenario for the evolution of complexity in animals, and indicate that inferences about the evolution of Metazoa based on the Ctenophora-sister hypothesis are not supported by the currently available data.

@@

Feuda *et al.* [@feuda2017improved] didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets). It was the first phylogenomic study that suggested recoding methods have better performance than non-recoding methods in the relationships at the root of the animal tree. The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (CAT-GTR) models in non-recoding and recoding datasets. The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with CAT-GTR models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).

Regarding the recovery of Porifera-sister, the authors concluded:

@@

### Whelan and Halanych 2016

Whelan *et al.* [@whelan2016let] is the only study to evaluate performance of site-heterogeneous models and site-homogeneous model with data partitioning under the simulation framework. The simulation results suggested that Poisson+CAT model consistently performed worse than other models in simulation datasets. More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimated substitutional heterogeneity in many cases. They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneous model. The results indicated that Poisson + CAT model tends to recover less accurate trees and more importantly,  both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.

The authors concluded:

@@

### Whelan *et al.* 2017

Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015, with the difference of a core Ctenophora core datasets were constructed here with more than 2000 genes. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in choanoazoa dataset. The main conclusions of this paper were based on Whelan2017_full and Whelan2017_strict matrices.

Regarding the recovery of Ctenophora-sister, the authors concluded:
>Using datasets with reasonably high ctenophore and other non-bilaterian taxon sampling, our results strongly reject the hypothesis that sponges are the sister lineage to all other extant metazoans.

@@

Simion *et al.* [@simion2017large] added transcriptomic data for 21 new animals. The data matrix was constructed using a semi-automated approach to comprehensively detect and eliminate potential systematic errors. The resulting dataset comprises 1,719 genes and 97 species, including 61 non-bilaterian species. It was by far the largest phylogenomic dataset in terms of taxon and gene sampling related to the relationship at the root of animal phylogeny.

The final matrix was first analyzed using the Poisson-CAT model. Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size. Each jackknife is baed on a random selection of ~ 25% of the genes. The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets. Importantly, the authors compared the behavior of long-branch effect between site-heterogeneous and site-homogeneous models by progressively removal taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.

@@

It should be noted that all the PhyloBayes runs have not reached convergence due to the computational cost in these large matrices.
